### 2021-04-14 pldi21/discussion.tex
cleanup spacing issues, add alt descriptions, misc cleanup
final content tweaks

\section{Discussion and Conclusion}\label{sec:discussion}\label{sec:conclusion}
\begin{quote}
  %
  \textit{
  %
  The arithmetical symbols are written diagrams and the geometrical figures are graphic formulas.
  %
  }

  \vspace{3pt}

@@

  The livelits mechanism as described in this paper operates only on expressions,
  but livelits might be useful for generating other sorts of terms, such as types,
  patterns, and entire modules. Prior work on literal macros has explored this \cite{TLMs}.
  (Note that Racket's visual macro system generates
  arbitrary syntax, so it can already be used in 
  this manner, albeit with no sort-specific semantic guarantees.)
  % Support for type and module splices, too, would likely be useful.
  % With these, better support for reflecting on the provided type might allow for 
  % the development of more general livelits, e.g. a variant of \li{\$datatype} 
  % with individually typed columns, rather than only floating point data.

>>>>

\section{Discussion and Conclusion}\label{sec:discussion}\label{sec:conclusion}
\begin{quote}
  %
  \textit{
  %
  The arithmetical symbols are written diagrams and the geometrical figures are graphic formulas.\vspace{3px}
  %
  }

@@

  The livelits mechanism as described in this paper operates only on expressions,
  but livelits might be useful for generating other sorts of terms, such as types,
  patterns, and entire modules. Prior work on literal macros has explored this \cite{TLMs}.
  Note that Racket's visual macro system generates
  arbitrary syntax, so it can already be used in 
  this manner, albeit with no sort-specific semantic guarantees.
  % Support for type and module splices, too, would likely be useful.
  % With these, better support for reflecting on the provided type might allow for 
  % the development of more general livelits, e.g. a variant of \li{\$datatype} 
  % with individually typed columns, rather than only floating point data.

### 2021-04-14 pldi21/implementation.tex
final content tweaks

Uniquely, every editor state in Hazel is 
semantically meaningful: it has a type, it can be evaluated, and it can be transformed 
in a type-aware manner. This implies that livelits remain fully functional at all times, 
even when the program is incomplete or erroneous. Hazel achieves this ``gap-free'' liveness guarantee by automatically inserting explicit holes as necessary 
while the user edits 
the program. Formally, Hazel is a type-aware structure editor \cite{Hazelnut}, rather than a text editor, 
although the developers are aiming to maintain a text-like experience (this effort is orthogonal to our own). 
To maintain this guarantee, Hazel inserts empty holes where there are missing terms, 
and non-empty holes as markers of errors. To maintain this invariant in the presence of livelits, 
new non-empty holes are needed to mark each of the failure modes suggested by the premises of the \rulename{ELivelit} rule in Fig.~\ref{fig:expansion}. 
In particular, non-empty holes mark (1) the invocation of an unbound livelit; (2) the invocation of a livelit 
with a model value of the wrong type; 
(3) a run-time error during evaluation of \li{init}, \li{update}, or \li{expand} functions (which manifests as a run-time hole); (4) 
expansion validation failure (i.e. that the parameterized expansion is of the wrong type given the splice list).  
Errors in view generation are not considered semantic errors (they display as error messages where the livelit GUI would have appeared.)

@@

\subsection{Text Editor Integration}
Livelits do not require the use of a structure editor. 
We have also developed a 
proof-of-concept implementation of livelit interaction in a textual program editor, Sketch-n-Sketch \cite{sns-pldi,sns-uist}.
Livelit interaction causes the serialized model in the text buffer to be changed, which updates the view.
This proof-of-concept is focused on this interaction loop so it is not at feature parity with the 
 main implementation, but it demonstrates that a syntax-recognizing text editor \cite{DBLP:journals/tosem/BallanceGV92,DBLP:conf/sde/HorganM84,interactive-visual-syntax} 
is sufficient to support livelits, albeit with some gaps in availability when 
there are syntax errors.
There are other systems that integrate visual syntax into an otherwise textual editor as well, notably Dr. Racket's recently introduced visual macro system \cite{interactive-visual-syntax}, discussed in the next section.

@@

\subsection{Integration into Imperative Languages}
\label{sec:imperative-langs}
Our focus in this paper was on pure languages.
Side effects pose a challenge on two fronts.
If they occur in a livelit implementation,
then the state exists at edit-time. 
The most viable approach would be to retain the model-view-update
architecture and specify that any transient mutable state will be reset between updates.
If side effects occur in livelit expansions, 
then the closure collection mechanism requires more 
care to avoid unsoundness, as described in Sec.~\ref{sec:calculus-closure-collection}.
% However, these problems are surmountable, and we look forward to 
% efforts to integrate livelits into imperative languages.

>>>>

Uniquely, every editor state in Hazel is 
semantically meaningful: it has a type, it can be evaluated, and it can be transformed 
in a type-aware manner. This implies that livelits remain fully functional at all times, 
even when the program is incomplete or erroneous. Hazel achieves this ``gap-free'' liveness guarantee by automatically inserting explicit holes as necessary 
while the user edits 
the program. Formally, Hazel is a type-aware structure editor \cite{Hazelnut}, rather than a text editor, 
although the developers are aiming to maintain a text-like experience (this effort is orthogonal to our own). 
To maintain this guarantee, Hazel inserts empty holes where there are missing terms, 
and non-empty holes as markers of errors. To maintain this invariant in the presence of livelits, 
new non-empty holes are needed to mark each of the failure modes suggested by the premises of the \rulename{ELivelit} rule in Fig.~\ref{fig:expansion}. 
In particular, non-empty holes mark (1) the invocation of an unbound livelit; (2) the invocation of a livelit 
with a model value of the wrong type; 
(3) a run-time error during evaluation of \li{init}, \li{update}, or \li{expand} functions (which would manifest as run-time holes); (4) 
expansion validation failure (i.e. that the parameterized expansion is of the wrong type given the splice list and expansion type).  
Errors in view generation are not considered semantic errors (they display as error messages where the livelit GUI would have appeared.)

@@

\subsection{Text Editor Integration}
Livelits do not require the use of a structure editor. 
We have also developed a 
proof-of-concept implementation of livelit interaction in a textual program editor, Sketch-n-Sketch \cite{sns-deuce}.
The livelit GUI appears in a pop up window when requested by the user. Interactions with this GUI cause the serialized model in the text buffer to be changed, which updates the view.
This proof-of-concept is not at feature parity with the Hazel  
 implementation, but it demonstrates that a syntax-recognizing text editor \cite{DBLP:journals/tosem/BallanceGV92,DBLP:conf/sde/HorganM84,interactive-visual-syntax} 
is sufficient to support livelits, albeit with gaps in availability when 
there are syntax errors.
There are other systems that integrate visual syntax into an otherwise textual editor as well, notably Dr. Racket's recent visual macro system \cite{interactive-visual-syntax}, discussed in the next section.

@@

\subsection{Integration into Imperative Languages}
\label{sec:imperative-langs}
Our focus in this paper was on pure languages.
Side effects pose a challenge on two fronts.

If they occur in a livelit implementation,
as is possible for livelits implemented using Hazel's support for Javascript-based 
livelits, 
then the state exists at edit-time. 
To ensure that the GUI's state persists and remains valid between reloads, 
it may then be necessary to persist 
the relevant portions of the edit-time state 
alongside the program, as in Smalltalk systems \cite{Goldberg84smalltalk-80-}, or else provide for a state re-initialization protocol
that is called when a program is re-loaded from persistent memory.
It may also be necessary to more carefully control the access that a livelit 
has to the editor itself: many participants in the Graphite survey indicated 
that they did not want palettes, which were able to make network requests at edit-time, 
to be able to access 
proprietary or sensitive source code \cite{Graphite}. 
In any case, the expansion should ideally be a pure function of the model even in an imperative language, 
or run as an isolated process each time, 
to ensure that program 
behavior does not depend on expansion-time editor state.

If side effects occur in livelit expansions, 
then the closure collection mechanism requires more 
care to avoid unsoundness, as described in Sec.~\ref{sec:calculus-closure-collection}.
In addition, continuous live evaluation as in Hazel is intrinsically problematic 
in an imperative setting because running code on each keystroke can cause arbitrary side
effects, e.g. network requests or I/O.

These problems are surmountable with suitable compromises, e.g. limiting continuous 
evaluation for code with external side effects, and we look forward to 
efforts to integrate livelits into imperative languages.

### 2021-04-14 pldi21/related-work.tex
final content tweaks

Recent work on a interactive visual macro system in Dr. Racket 
is quite similar in spirit to our work \cite{interactive-visual-syntax}.
However, it supports only a limited form of splicing via a pop up text editor
that does not support full compositionality as described here, where livelits 
can appear within other livelits. Splicing is not strictly hygienic, though 
Racket's scope management machinery can be used by macros to voluntarily 
maintain a binding discipline \cite{DBLP:conf/popl/Flatt16}. There is not any 
consideration of typing. Parameterization and partial application is also impossible, because invocation 
is via an editor command rather than a syntactic mechanism. Finally,
there is no support for liveness: macro evaluation occurs in the editor 
environment, not the run-time environment of the program being written, 
so splice evaluation would not work correctly (except in trivial cases
where the splice was, e.g., a closed expression, or used only standard 
library constructs). That said, we conjecture that it is
likely possible to implement many of the mechanisms we describe in this paper 
perhaps in some form as a layer atop existing mechanisms in Dr. Racket.

Our structural delimitation of splices is reminiscent of work on 
\emph{language boxes} \cite{DBLP:conf/sle/RenggliDN09}, which focused on 
combining different notations, primarily textual, using structural delimiters inserted 
explicitly using a special-purpose editor.

There has been a long line of research on \emph{projectional editing}, where
the user edits graphical representations (projections) of code constructs 
\cite{DBLP:journals/ile/MillerPMV94,read1996generating,DBLP:conf/sde/Reiss84,DBLP:conf/uist/KoM05}. Livelits are a form of projectional editing. 
Many of the oldest systems offer only a fixed set of projections and interaction techniques. 
More recently, language workbenches with support for projectional editing like Citrus \cite{DBLP:conf/uist/KoM05}
and MPS \cite{voelter2011language} have made it easier to define new projectional editors.
However, these systems generate entire editors, whereas livelit definitions 
are decentralized in libraries. 
Furthermore, livelit definitions are governed by a type and binding discipline.
% In many ways, an unfamiliar livelit can be approached much like an unfamiliar function
% can be approached: by inspecting its declarations and reasoning compositionally 
% about arguments, i.e. splices and parameters.
Finally, livelits are uniquely live, building on Hazelnut Live \cite{HazelnutLive}.

@@

Related to projectional editors are 
a variety of systems that generate visualizations from code \cite{DBLP:conf/chi/Lerner20,koschke2003software,urquiza2004survey}. Livelits differ 
from these systems in directionality: visualization systems are given code and generate
a visualization, whereas livelits generate code where there would otherwise
be a hole.

Conal Elliott's work on tangible functional programming \cite{DBLP:conf/icfp/Elliott07} similarly explored a
system allows editing a graphical representation of code in compositional ways, 
but the editing representation itself is fixed as a series of connected windows.
Only the visual representation is customizable.

>>>>

This structural delimitation of splices is reminiscent of work on 
\emph{language boxes} \cite{DBLP:conf/sle/RenggliDN09}, which focused on 
combining different notations, primarily textual, using structural delimiters inserted 
explicitly using a special-purpose editor.

Recent work on a interactive visual macro system in Dr. Racket 
is quite similar in spirit to our work \cite{interactive-visual-syntax}.
However, it supports only a limited form of splicing via a pop up text editor
that does not support full compositionality as described here, where livelits 
can appear within other livelits. Splicing is not strictly hygienic, though 
Racket's scope management machinery can be used by macros to voluntarily 
maintain a binding discipline \cite{DBLP:conf/popl/Flatt16}. There is not any 
consideration of typing. Parameterization and partial application is also impossible, because invocation 
is via an editor command rather than a syntactic mechanism. Finally,
there is no support for liveness: macro evaluation occurs in the editor 
environment, not the run-time environment of the program being written, 
so splice evaluation would not work correctly (except in trivial cases
where the splice was, e.g., a closed expression, or used only standard 
library constructs). That said, we conjecture that it is
possible to implement many of the mechanisms we describe in this paper 
in some form as a layer atop these existing mechanisms in Dr. Racket.

There has been a long line of research on \emph{projectional editing}, where
the user edits graphical representations (projections) of code constructs 
\cite{DBLP:journals/ile/MillerPMV94,read1996generating,DBLP:conf/sde/Reiss84,DBLP:conf/uist/KoM05}. Livelits are a form of projectional editing. 
Many of the oldest systems offer only a fixed set of projections and interaction techniques. 
More recently, language workbenches with support for projectional editing like Citrus \cite{DBLP:conf/uist/KoM05}
and MPS \cite{voelter2011language} have made it easier to define new projectional editors.
However, these systems generate entire editors, whereas livelit definitions 
are decentralized in libraries and lexically scoped. 
Furthermore, livelit definitions are governed by a type and binding discipline.
% In many ways, an unfamiliar livelit can be approached much like an unfamiliar function
% can be approached: by inspecting its declarations and reasoning compositionally 
% about arguments, i.e. splices and parameters.
Finally, livelits are uniquely live, building on hole closure tracking 
from Hazelnut Live \cite{HazelnutLive}.

@@

Related to projectional editors are 
a variety of systems that generate visualizations from code \cite{DBLP:conf/chi/Lerner20,koschke2003software,urquiza2004survey,Goldberg84smalltalk-80-}. Livelits differ 
from these systems in directionality: visualization systems are given code and generate
a visualization, whereas livelits generate code where there would otherwise
be a hole.

Conal Elliott's work on tangible functional programming \cite{DBLP:conf/icfp/Elliott07} similarly explored a
system allows editing a graphical representation of code in compositional ways, 
but the editing representation itself is fixed as a series of connected windows.
Only the visualizations are customizable.
